Msprime 1.0: an efficient and extensible coalescent simulation framework

### 2021-07-23 paper.tex
rearranging mutation section

\subsection*{Simulating mutations}
\label{sec-mutations}
% Explain the background: what have people done classically?
Because coalescent simulations are usually concerned with
neutral evolution (see the \nameref{sec-selection} section, however)
the problem of generating synthetic genetic variation can be decomposed into
two independent steps:
simulating genetic ancestry (the trees), and then simulating
variation by superimposing mutation processes on those trees.
The classical \texttt{Seq-Gen} program~\citep{rambaut1997seq}
allows a range of different models of sequence evolution to be imposed on a given
tree in Newick format, and there are various
extensions to the basic nucleotide
models~\citep[e.g.][]{cartwright2005dna,fletcher2009indelible}.
Population genetic simulators have tended to
include their own implementations of mutation simulation, with
most just supporting the infinite sites
model~\citep[e.g.][]{hudson2002generating}
but with several supporting a range of different models
of sequence
evolution~\citep[e.g.][]{mailund2005coasim,excoffier2011fastsimcoal,
virgoulay2021gspace}.
% Mutations are not treated as first-class data --- the sequences
% were really all people were after, classically.
The output of the combined simulations of ancestry and mutations is
then some representation of the resulting nucleotide sequences
(or some derived statistics of those sequences);
information about the mutations that have been simulated
is usually not retained or made available for subsequent analysis.

The approach taken in \msprime\ is quite different.
As illustrated in Fig.~\ref{fig-ts-example}, mutations are an
integral part of \tskit's tree sequence data model, and genetic
variation is encoded by sites and the mutations at those sites;
the sequences themselves are never stored.
This may seem inconvenient at first, but it is extremely powerful.
Firstly, the storage space required for simulations is dramatically
reduced. For a simulation of $n$ samples with $m$ variant sites,
we required $O(nm)$ space to store the sequence data as a variant
matrix. However, if this simulation was of a recombining genome
with $t$ trees, then the \tskit\ tree sequence encoding
requires $O(n + t + m)$ space~\citep{kelleher2016efficient},
assuming we have $O(1)$ mutations at each site. For
large sample sizes, this difference is profound, making it
conceivable, for example, to store the genetic ancestry
and variation data for the entire human population on a
laptop~\citep{kelleher2019inferring}. Besides the huge
difference in storage efficiency, it is also often far more efficient
to compute statistics of the sequence data from the
trees and mutations than it is to work with the sequences themselves.
For example, computing Tajima's $D$ from simulated data using \tskit\
is several orders of magnitude faster than efficient variant matrix
libraries for large sample sizes~\citep{ralph2020efficiently}.

The advantages of storing mutations rather than sequences have
recently become clear in the context of the vast genomic
datasets produced during the SARS-CoV-2 pandemic.
\cite{turakhia2021ultrafast} propose the
Mutation Annotated Tree (MAT)
format (consisting of a Newick tree
and associated mutations in a binary format) and
the \texttt{matUtils} program as an efficient way to store
and process large viral datasets~\citep{mcbroome2021daily},
achieving excellent compression and processing performance.
Similarly, \phastsim~\citep{demaio2021phastsim}
was developed to simulate sequence evolution on such large SARS-CoV-2
phylogenies, and also outputs a Newick tree annotated with mutations (not
in MAT format) to avoid the bottleneck of generating
and storing the simulated sequences.
While these methods illustrate the advantages of the general
approach of storing ancestry and mutations rather than sequences,
they do not generalise beyond their immediate settings, and
no software library support is available.

% on the other hand, tskit is awesome.
The \tskit\ library provides first-class support for site and mutation
data, fully integrated with the topological information encoded in the
node and edge tables (Fig~\ref{fig-ts-example}). The mature and
efficient C and Python interfaces makes it straightforward to add
further information to an existing file, and to output the results.
As such, it is essentially a historical accident
that \msprime\ (whose main focus is on ancestry simulation)
contains code to simulate mutations: the process could (and arguably should)
be performed in a separate package, and there would be no performance
penalty for doing so. It is important to note that \msprime's
mutation generator can work with \emph{any} \tskit\ tree sequence,
be it simulated using \SLiM~\citep{haller2019slim} or
\fwdpy~\citep{thornton2014cpp}, or estimated from real data
using~\texttt{tsinfer}/\texttt{tsdate}~\citep{kelleher2019inferring,wohns2021unified}
or \texttt{Relate}~\citep{speidel2019method}. It is a modular
component intended to fit into a larger software ecosystem, and
is in no way dependent on \msprime's ancestry simulator.

For version 1.0 we have greatly extended the sophistication of
\msprime's mutation generation engine, achieving near feature-parity
with \SeqGen.
We  support a large number of mutation models, including
JC69~\citep{jukes1969evolution},
F84~\citep{felsenstein1996hidden},
and GTR~\citep{tavare1986some} nucleotide models
and the BLOSUM62~\citep{henikoff1992amino}
and PAM~\citep{dayhoff1978} amino acid models.
The general model accepts an arbitrary transition matrix between
any number of alleles (which can be arbitrary unicode strings),
and so other models such as the Kimura two and three
parameter models~\citep{kimura1980simple,kimura1981estimation}
can be easily and efficiently defined in user code.
Mutation rates can vary along the genome, and multiple mutation
models can be imposed on a tree sequence by overlaying mutations
in multiple passes.
The new \texttt{sim\_mutations} function
(see the \nameref{sec-sim-interface} section) defaults to a finite-sites
mutation model, but full support for the
original infinite sites model is also maintained.
We have extensively validated the results of mutation simulations
against both theoretical expectations and output from
\SeqGen~\citep{rambaut1997seq} and
Pyvolve~\citep{spielman2015pyvolve}.

>>>>

\subsection*{Genetic variation}
\label{sec-genetic-variation}

The output of population genetics simulators, quite naturally,
tends to be some representation of the resulting nucleotide sequences
(or some derived statistics of those sequences),
although the underlying genealogies are also of interest.
For instance,
the classic simulator \ms\ will output trees or genome sequences,
but not both.
% Mutations are not treated as first-class data --- the sequences
% were really all people were after, classically.
However, joint information about the mutations that have been simulated
-- e.g., where they have occurred on the trees -- 
is usually not retained or made available for subsequent analysis.

The approach taken in \msprime\ is quite different.
As illustrated in Fig.~\ref{fig-ts-example}, mutations are an
integral part of \tskit's tree sequence data model, and genetic
variation is encoded by recording sites at which mutations
have occurred, and where each mutation at those sites
has occurred on the marginal tree;
the sequences themselves are never stored.
It may at first seem inconvenient to have only this indirect representation
of the genome sequences, but it is extremely powerful.
Firstly, the storage space required for simulations is dramatically
reduced. For a simulation of $n$ samples with $m$ variant sites,
we would require $O(nm)$ space to store the sequence data as a variant
matrix. However, if this simulation was of a recombining genome
with $t$ trees, then the \tskit\ tree sequence encoding
requires $O(n + t + m)$ space~\citep{kelleher2016efficient},
assuming we have $O(1)$ mutations at each site. For
large sample sizes, this difference is profound, making it
conceivable, for example, to store the genetic ancestry
and variation data for the entire human population on a
laptop~\citep{kelleher2019inferring}. Besides the huge
difference in storage efficiency, it is also often far more efficient
to compute statistics of the sequence data from the
trees and mutations than it is to work with the sequences themselves.
For example, computing Tajima's $D$ from simulated data using \tskit\
is several orders of magnitude faster than efficient variant matrix
libraries for large sample sizes~\citep{ralph2020efficiently}.

The advantages of storing genetic variation data using the underlying trees
have recently become clear in the context of the vast genomic
datasets produced during the SARS-CoV-2 pandemic.
\cite{turakhia2021ultrafast} propose the
Mutation Annotated Tree (MAT)
format (consisting of a Newick tree
and associated mutations in a binary format) and
the \texttt{matUtils} program as an efficient way to store
and process large viral datasets~\citep{mcbroome2021daily},
achieving excellent compression and processing performance.
Similarly, \phastsim~\citep{demaio2021phastsim}
was developed to simulate sequence evolution on such large SARS-CoV-2
phylogenies, and also outputs a Newick tree annotated with mutations (not
in MAT format) to avoid the bottleneck of generating
and storing the simulated sequences.
While these methods illustrate the advantages of the general
approach of storing ancestry and mutations rather than sequences,
they do not generalise beyond their immediate settings, and
no software library support is available.
\msprime, on the other hand, provides a fast, powerful, and well-tested
generic interface for generating and storing mutations on trees.

\subsection*{Simulating mutations}
\label{sec-mutations}

% Explain the background: what have people done classically?
Because coalescent simulations are usually concerned with
neutral evolution (see the \nameref{sec-selection} section, however)
the problem of generating synthetic genetic variation can be decomposed into
two independent steps:
simulating genetic ancestry (the trees), and then simulating
variation by superimposing mutation processes on those trees.
Indeed, to call a mutation ``neutral'' means that its presence or absence
does not affect demography.
A number of programs exist to place mutations on trees:
in addition to \phastsim~\citep{demaio2021phastsim},
the classical \texttt{Seq-Gen} program~\citep{rambaut1997seq}
allows a range of different models of sequence evolution to be imposed on a given
tree in Newick format, and there are various
extensions to the basic nucleotide
models~\citep[e.g.][]{cartwright2005dna,fletcher2009indelible}.
Population genetic simulators have tended to
include their own implementations of mutation simulation, with
most just supporting the infinite sites
model~\citep[e.g.][]{hudson2002generating}
but with several supporting a range of different models
of sequence
evolution~\citep[e.g.][]{mailund2005coasim,excoffier2011fastsimcoal,
virgoulay2021gspace}.

% on the other hand, tskit is awesome.
Clean separation of the ancestry and mutation generation steps
is made possible by the \tskit\ library, whose mature and
efficient C and Python interfaces makes it straightforward to add
further information to an existing file, and to output the results.
As such, it is essentially a historical accident
that \msprime\ (whose main focus is on ancestry simulation)
contains code to simulate mutations: the mutation process could
be performed by a separate package, and there would be no performance
penalty for doing so.
Thanks to this interoperability, \msprime's
mutation generator can work with \emph{any} \tskit\ tree sequence,
be it simulated using \SLiM~\citep{haller2019slim} or
\fwdpy~\citep{thornton2014cpp}, or estimated from real data
using~\texttt{tsinfer}/\texttt{tsdate}~\citep{kelleher2019inferring,wohns2021unified}
or \texttt{Relate}~\citep{speidel2019method}. It is a modular
component intended to fit into a larger software ecosystem, and
is in no way dependent on \msprime's ancestry simulator.

For version 1.0 we have greatly extended the sophistication of
\msprime's mutation generation engine, achieving near feature-parity
with \SeqGen.
We  support a large number of mutation models, including
JC69~\citep{jukes1969evolution},
F84~\citep{felsenstein1996hidden},
and GTR~\citep{tavare1986some} nucleotide models
and the BLOSUM62~\citep{henikoff1992amino}
and PAM~\citep{dayhoff1978} amino acid models.
A general model accepts an arbitrary transition matrix between
any number of alleles (which can be arbitrary unicode strings),
and so other models such as the Kimura two and three
parameter models~\citep{kimura1980simple,kimura1981estimation}
can be easily and efficiently defined in user code.
Mutation rates can vary along the genome, and multiple mutation
models can be imposed on a tree sequence by overlaying mutations
in multiple passes.
The new \texttt{sim\_mutations} function
(see the \nameref{sec-sim-interface} section) defaults to a finite-sites
mutation model, but full support for the
original infinite sites model is also maintained.
We have extensively validated the results of mutation simulations
against both theoretical expectations and output from
\SeqGen~\citep{rambaut1997seq} and
Pyvolve~\citep{spielman2015pyvolve}.
